---
title: "Risk_model_Code_and_description"
author: "Courtney Schreiner"
date: "2024-10-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step by step for creating risk model code

For this document I hope to go through my code for the risk model and double check that each piece does what it is supposed to do. The main document has functions first and then parameter declaration and then running the simulations. For this document I will go in order of the pieces needed, so I will start withe the community level model, then to adjacency matrix, transition, and then lastly all the code need to solve the system of equations

```{r packages}

library(deSolve)
library(tidyverse)
library(igraph)
library(ggnetwork)
library(GGally)
library(intergraph)
library(ggthemes)
library(paletteer)
library(pals)
library(ggrepel)
library(ggpubr)

```

  
  **Community model parameters**
```{r Community model parameters}
community_pop <- 100000 #pop size of the larger community
init_conds <- c(S = community_pop-1, I = 1, R = 0) 
print(init_conds)
parms <- data.frame(bet =0.0000035 , gam = 1/21) 
print(parms)
times <- seq(from = 1, to = 144, by = 1)
print(times)


```
**Community modelfunction/equations**
```{r communtiy model function}
SIR_community_model <- function(t, x, parms) {
  S <- x[1]
  I <- x[2]
  R <- x[3]
  
  with(parms, {
    dS <- -bet*S*I
    dI <- bet*S*I - gam*I
    dR <- gam*I
    
    dt <- c(dS,dI,dR)
    return(list(dt))
  })}

```


**Community model output**

```{r pressure, echo=FALSE}

Community_output <- data.frame(lsoda(y = init_conds, func = SIR_community_model,times = times, parms=parms))

Community_output %>% pivot_longer(cols = !time) %>% arrange(desc(time))%>% 
  ggplot(aes(x=time,y =value, color = name))+geom_line()+
  scale_color_manual(name=NULL,values=c("blue","red","purple"),breaks = c("S","I","R"))+
  theme_classic()+
  labs(x= "Time (days)", y = "Number of individuals")+ggtitle("Community outbreak of pathogen")

```

**Make building**
```{r}
church_adjacency_matrix <- matrix(c(c(0,1,1,1,1),
                                    c(1,0,1,1,0),
                                    c(1,1,0,1,0),
                                    c(1,1,1,0,0),
                                    c(1,0,0,0,0)),nrow=5, ncol = 5)
church_graph<- graph_from_adjacency_matrix(church_adjacency_matrix, mode = "undirected")
#take a look
plot(church_graph)

```
```{r set carrying capacities}
graph_to_use <- church_graph
adjacency_matrix_to_use <- church_adjacency_matrix 
N_rooms <- ncol(adjacency_matrix_to_use) #number of rooms
Church_C<- c(200, #column 1 - main area / Hallway
       100, # column 2 - Hallway
       100, # column 3 - Hallway
       400, #column 4 Main room
       200
) 

length(Church_C)
```
Now that we have building defined by the adjacency matrix, we need to specify the conditions in the building
```{r Building paramaters}
Max_Building_Capacity <- sum(Church_C) 
Prop_full <- 0.8 # how full do we want our building capacity to be
Adj_Max_Building_Capacity <- Prop_full*Max_Building_Capacity #Adjusted capacity - number of individuals that will be in the building such that building is at 'Prop_full' capacity
delt <- Adj_Max_Building_Capacity/community_pop # proportionality constant ( what proportion of individuals from the community are in the building of interest)

```

Next, we need to put people in our building. These will be our initial conditions for the building level model.

  
```{r initialize building}

day <- 31 #What day to extract results from the community level model
N_rooms <- length(Church_C)
Bld_setup_func <- function(Community_output,day, delt,N_rooms){
  Building_ICs <- Community_output[day,] #Retrieves the number of S, I, and R individuals in the community at a particular day.
  
  #S_c + I_c + R_c = N_c | sum of the number of susceptible, infectious and recovered in the community is the total population of the community
  # we want the proportion of S, I, and R in the community to match the proportion of S, I, and R in the building:
  # (S_c + I_c + R_c)*delt = building_pop
  # we set building_pop based on how full we want the building to be: this is our adjusted_building capacity set above
  #i.e. Prop_Full*Building_Max_Capacity= Adjusted_building_capacity.
  # then building_pop = Adjusted_building_capacity so that we can get our desired amount of individuals in the building. 
  # and then we solve for delt above: delt = adjusted_building_pop/community_pop 
  
  Sb <- Community_output$S[day]*delt # number of susceptible in building
  Ib <- Community_output$I[day]*delt # number of Infectious in building
  Rb <- Community_output$R[day]*delt # number of Recovered in building
  
  #Sb, Ib, Rb are the number of Susceptible, Infected and Recovered individuals that will be in the building
  
  #initial conditions for each room. Randomly distribute individuals throughout rooms
  
  # first assign a random number between 0 and 1 for each room,
  #broken up by S, I, and R
  S_x <- c(runif(N_rooms, min = 0, max = 1))
  I_x <- c(runif(N_rooms, min = 0, max = 1))
  R_x <- c(runif(N_rooms, min = 0, max = 1))
  
  #normalize and then assign the correct number of S, I, and R based on Sb, Ib, and Rb
  S_x <- ((S_x/sum(S_x))*Sb)
  I_x <- ((I_x/sum(I_x))*Ib)
  R_x <- ((R_x/sum(R_x))*Rb)
  
  
  S_prop <- ((S_x/sum(S_x))*Sb)/(Sb+Ib+Rb)
  I_prop <- ((I_x/sum(I_x))*Ib)/(Sb+Ib+Rb)
  R_prop <- ((R_x/sum(R_x))*Rb)/(Sb+Ib+Rb)
  
  #this model is of the proportion of S, I and R in each room so now we need to switch from numbers to proportions.
  # but we need to keep the total number of individuals in each room since that is in our model
  Init_conds_nums <- data.frame(S_num = S_x, I_num = I_x, R_num = R_x)
  Init_conds_nums <- Init_conds_nums %>% mutate(N_x = S_num + I_num + R_num)
  Init_conds_props <- Init_conds_nums %>% 
    mutate(S_prop = S_num/N_x, I_prop = I_num/N_x, R_prop = R_num/N_x)%>%
    select(S_prop,I_prop,R_prop,N_x)
  
  # we start with no particles in the building
  P_x <- c(rep(0,N_rooms))
  
  Init_conds_props <- Init_conds_props %>% mutate(S_prop = S_prop,I_prop =I_prop,R_prop= R_prop, P_x =P_x,N_x=N_x)
  #Init_conds <-c(S=S_x/sum(Sb+Ib+Rb), I = I_x/sum(Sb+Ib+Rb), R = R_x/sum(Sb+Ib+Rb), P = P_x)
  
  return(data.frame(S=S_prop,I=I_prop,R=R_prop, P=P_x, N_x = S_x+I_x+R_x))
}

Church_setup <- Bld_setup_func(Community_output = Community_output,day = day,delt = delt,N_rooms =N_rooms)

```
**Checking initial conditions**
Now lets check that our building setup function did what I wanted it to. 
Each room's proportion should sum to 1. And if we convert back to numbers using N_x the numbers should equal
```{r props sum to 1}
#Church_setup %>% mutate(total_prop = S_prop+I_prop+R_prop)
```
props sum to 1 so that is good

```{r props in building match community}
#prop of susceptible in building
sum(Church_setup$S_prop*Church_setup$N_x)/sum(Church_setup$N_x)
#prop of susceptible in community
Community_output$S[day]/(Community_output$S[day]+Community_output$I[day]+Community_output$R[day])

#prop of infectious in building
sum(Church_setup$I_prop*Church_setup$N_x)/sum(Church_setup$N_x)
#prop of infectious in community
Community_output$I[day]/(Community_output$S[day]+Community_output$I[day]+Community_output$R[day])

#prop of recovered in building
sum(Church_setup$R_prop*Church_setup$N_x)/sum(Church_setup$N_x)
#prop of susceptible in community
Community_output$R[day]/(Community_output$S[day]+Community_output$I[day]+Community_output$R[day])


```
Okay, great all of that checks out

**Now people and particles need to move (Transition matrices)**

```{r}
Create_T_Matrix <-function(adjacency_matrix_to_use){
  #set.seed(123145) # <- easier for debugging
  T_mov <- data.frame(matrix(runif(N_rooms^2), nrow = N_rooms)) #populates a square matrix/dataframe with random numbers between 0 and 1 for the number of rooms that our building has.
  diag(T_mov) <- 0 #set diagonal to 0
  T_mov <- adjacency_matrix_to_use*T_mov #restrict the movement according to our network/adjacency matrix
  T_mov_norm <- t(apply(T_mov, 1, function(x) x / sum(x))) # normalize so that there aren't more people moving than what can (rows should sum to 1)
  T_mov <-T_mov_norm
  return(T_mov)
}

Church_T_mov <- Create_T_Matrix(adjacency_matrix_to_use = adjacency_matrix_to_use)
sum(Church_T_mov[1,])
sum(Church_T_mov[2,])
sum(Church_T_mov[3,])
sum(Church_T_mov[4,])
sum(Church_T_mov[5,])

# We can use the same function for the particle matrix
Church_theta_mov <- Create_T_Matrix(adjacency_matrix_to_use = adjacency_matrix_to_use)

```

**Now set parameters for the model**


```{r parameters}
parms <-data.frame(s=100,a=5, d=3,lam = 1)
# s = shedding, a = absorption, d = decay, lam = scalar for room capacities

Maxtime <- 24*3
times <- seq(from = 0, to = Maxtime, by = 0.2)
m <-5 #number of equations per room
```

For the model we are going to have to sum over some of the vectors/matrices so we will put those steps in functions. This is needed for the movement of individuals and particles. Particles and people will have to have a different function because there re capacities on rooms for people but not particles.

```{r movement functions}

flux_in_people <- function(N_rooms, Transition_matrix,State,Room_pops,Carrying_capacity,t){
  # Transition_matrix <- matrix(c(0,0.5,1,0.7,0,0,0.3,0.5,0), nrow = 3,ncol = 3)
  # State <- Church_setup$S_prop
  # Room_pops <- Church_setup$N_x
  # Carrying_capacity <- Church_C
  all_room_change <- c(seq(N_rooms))
  for(x in 1:N_rooms){ #flow in to room x from other rooms (j)
    flux_in_temp <- 0
    for(j in 1:N_rooms){
      flux_in_temp <- flux_in_temp + State[j]*Transition_matrix[j,x]*(1-(Room_pops[x]/Carrying_capacity[x]))
    }
    all_room_change[x] <- flux_in_temp
  }
  
  test <- c(M = as.vector(all_room_change))
  return(test)
}

flux_out_people <- function(N_rooms, Transition_matrix,State,Room_pops,Carrying_capacity){
  all_room_change <- c(seq(N_rooms))
  for(x in 1:N_rooms){
    flux_out_temp <- 0
    for(j in 1:N_rooms){
      flux_out_temp <- flux_out_temp + State[x]*Transition_matrix[x,j]*(1-(Room_pops[j]/Carrying_capacity[j]))
    }
    all_room_change[x] <- flux_out_temp
  }
  test <- c(M = as.vector(all_room_change))
  return(test)
}

flux_in_particles <- function(N_rooms, Transition_matrix,State){
  all_room_change <- c(seq(N_rooms))
  for(x in 1:N_rooms){
    flux_in_temp <- 0
    for(i in 1:N_rooms){
      flux_in_temp <- flux_in_temp + State[i]*Transition_matrix[i,x]
    }
    all_room_change[x] <- flux_in_temp
  }
  
  test <- c(M = as.vector(all_room_change))
  return(test)
}

flux_out_particles <- function(N_rooms, Transition_matrix,State){
  all_room_change <- c(seq(N_rooms))
  for(x in 1:N_rooms){
    flux_out_temp <- 0
    for(i in 1:N_rooms){
      flux_out_temp <- flux_out_temp + State[x]*Transition_matrix[x,i]
    }
    all_room_change[x] <- flux_out_temp
  }
  test <- c(M = as.vector(all_room_change))
  return(test)
}

```


**Function for the whole model**
```{r function for model}
Particle_model <- function(t, x, parms,T_mov, theta_mov, adjacency_matrix_to_use,C_x){
  ncompartment <- 5
  n_rooms <- length(x)/ncompartment
  S <- as.matrix(x[1:n_rooms])
  I <- as.matrix(x[(n_rooms+1):(2*n_rooms)])
  R <- as.matrix(x[(2*n_rooms+1):(3*n_rooms)])
  P <- as.matrix(x[(3*n_rooms+1):(4*n_rooms)])
  N_x <- as.matrix(x[(4*n_rooms+1):(5*n_rooms)])

  with(parms,{
  
    
    dS <- as.matrix((flux_in_people(N_rooms, Transition_matrix =T_mov, State=S,Room_pops = N_x,Carrying_capacity = C_x)) - flux_out_people(N_rooms, Transition_matrix = T_mov, State=S,Room_pops = N_x,Carrying_capacity = C_x))
    
    dI <- as.matrix((flux_in_people(N_rooms, Transition_matrix =T_mov, State=I,Room_pops = N_x,Carrying_capacity = C_x)) - flux_out_people(N_rooms, Transition_matrix = T_mov, State=I,Room_pops = N_x,Carrying_capacity = C_x))
    
    dR <- as.matrix((flux_in_people(N_rooms, Transition_matrix =T_mov, State=R,Room_pops = N_x,Carrying_capacity = C_x)) - flux_out_people(N_rooms, Transition_matrix = T_mov, State=R,Room_pops = N_x,Carrying_capacity = C_x))
    #last step will be the particle EQ
    
    dP <- s*as.matrix(I)*as.matrix(N_x) - as.matrix(a*P/(lam*C_x*N_x))-d*as.matrix(P) + as.matrix(as.matrix(flux_in_particles(N_rooms, theta_mov,State = P)) - as.matrix(flux_out_particles(N_rooms, theta_mov,State = P)))
    
    dN_x <- as.matrix((flux_in_people(N_rooms, Transition_matrix =T_mov, State=N_x,Room_pops = N_x,Carrying_capacity = C_x)) - flux_out_people(N_rooms, Transition_matrix = T_mov, State=N_x,Room_pops = N_x,Carrying_capacity = C_x))
    
    dt <- c(dS,dI,dR,dP,dN_x)
    return(list(dt))})
  
}
```


```{r}

Church_Init_conds <-c(S=Church_setup$S, I = Church_setup$I, R = Church_setup$R, P = Church_setup$P, N_x =Church_setup$N_x)

Church_output <- data.frame(lsoda(y = Church_Init_conds, func = Particle_model,times = times,
                                  parms = parms,
                                  adjacency_matrix_to_use=adjacency_matrix_to_use,
                                  theta_mov =Church_theta_mov,
                                  T_mov = Church_T_mov, 
                                  C_x=Church_C))
```


```{r}
Church_data_clean <- Church_output%>% pivot_longer(cols = !time,
                                                   names_to = c("State", "Room"),
                                                   names_pattern = "([A-Za-z]+)(\\d+)",
                                                   values_to = "Number")

# Church_data_ratios <- Church_data_clean %>% pivot_wider(names_from = c(State),values_from = c(Number)) %>% group_by(time,Room)  %>% 
#   mutate(N_x = S+I+R) %>% group_by(time,Room) %>% 
#   mutate(K_x = ((parms$s)*I)/((parms$a)*N_x), prop_to_K = P/K_x)
# Church_P_x_K_x_plot<-ggplot(Church_data_ratios,aes(x =time, y =prop_to_K,group= Room,color=Room))+geom_line()+labs(title = "Church - Risk - 31 rooms")
# Church_P_x_K_x_plot
# Church_P_x_plot <- ggplot(Church_data_ratios, aes(x=time, y = P, color = Room))+geom_line()+labs(title = "Church - Risk - 31 rooms")
# Church_P_x_plot

Church_data_clean %>% filter(State == "S") %>% 
  group_by(State,Room)%>% 
  ggplot( aes(x=time, y = Number, group =Room, color = Room))+
  geom_line()+theme_classic()+
  labs(x = "Time (hours)", y= "Proportion of Susceptible individuals")+ggtitle("Proportion of Susceptible individuals across rooms")

Church_data_clean %>% filter(State == "I") %>% 
  group_by(State,Room)%>% 
  ggplot( aes(x=time, y = Number, group =Room, color = Room))+
  geom_line()+theme_classic()+
  labs(x = "Time (hours)", y= "Proportion of Infectious individuals")+ggtitle("Proportion of Infectious individuals across rooms")
Church_data_clean %>% filter(State == "R") %>% 
  group_by(State,Room)%>% 
  ggplot( aes(x=time, y = Number, group =Room, color = Room))+
  geom_line()+theme_classic()+
  labs(x = "Time (hours)", y= "Proportion of Recovered individuals")+ggtitle("Proportion of Recovered individuals across rooms")
Church_data_clean %>% filter(State == "P") %>% 
  group_by(State,Room)%>% 
  ggplot( aes(x=time, y = Number, group =Room, color = Room))+
  geom_line()+theme_classic()+
  labs(x = "Time (hours)", y= "Number of infectious particles")+ggtitle("Infectious particles in rooms within a building")
Church_data_clean %>% filter(State == "x") %>% 
  group_by(State,Room)%>% 
  ggplot( aes(x=time, y = Number, group =Room, color = Room))+
  geom_line()+theme_classic()+
  labs(x = "Time (hours)", y= "Number of individuals")+ggtitle("Number of individuals in rooms within a building")

# Church_data_clean %>% filter(State != "P") %>% 
#   group_by(time,Room)%>% summarise(total_room_pop=sum(Number)) %>% group_by(time) %>% summarise(tot_bld_pop =sum(total_room_pop)) %>% 
#   ggplot( aes(x=time, y = tot_bld_pop))+
#   geom_line()+theme_classic()+
#   labs(x = "Time (hours)", y= "Number of people")+ggtitle("number of people in building")

Church_data_clean %>% filter(State == "S"| State == "I" | State == "R") %>% ungroup() %>% group_by(time, Room) %>% summarise(Total_prop=sum(Number)) %>%
  ggplot(aes(x=time, y=Total_prop, color=Room))+geom_line()+
  labs(x = "Time (hours)", y= "Proportion of individuals")+ggtitle("Proportion of individuals in rooms within a building")

Church_data_clean %>% filter(State == "S"| State == "I" | State == "R") %>% ungroup() %>% group_by(time) %>% summarise(Total_prop=sum(Number)) %>%
  ggplot(aes(x=time, y=Total_prop))+geom_line()+
  labs(x = "Time (hours)", y= "Proportion of individuals")+ggtitle("Proportion of individuals in the building")


Church_data_clean %>% filter(State== "x")%>% ungroup() %>% group_by(time) %>% summarise(Total_pop=sum(Number)) %>%
  ggplot(aes(x=time, y=Total_pop))+geom_line()+
  labs(x = "Time (hours)", y= "Number of individuals")+ggtitle("Number of individuals in the building")

```

