---
title: "Risk_model_Code_and_description"
author: "Courtney Schreiner"
date: "2024-10-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step by step for creating risk model code


```{r packages, include=FALSE}

library(deSolve)
library(tidyverse)
library(igraph)
library(ggnetwork)
library(GGally)
library(intergraph)
library(ggthemes)
library(paletteer)
library(pals)
library(ggrepel)
library(ggpubr)

```

  
  **Community model parameters**
```{r Community model parameters}
community_pop <- 100000 #pop size of the larger community
init_conds <- c(S = community_pop-1, I = 1, R = 0) 
print(init_conds)
parms <- data.frame(bet =0.0000035 , gam = 1/21) 
print(parms)
times <- seq(from = 1, to = 144, by = 1)
print(times)


```
**Community modelfunction/equations**
```{r communtiy model function}
SIR_community_model <- function(t, x, parms) {
  S <- x[1]
  I <- x[2]
  R <- x[3]
  
  with(parms, {
    dS <- -bet*S*I
    dI <- bet*S*I - gam*I
    dR <- gam*I
    
    dt <- c(dS,dI,dR)
    return(list(dt))
  })}

```


**Community model output**

```{r pressure, echo=FALSE}

Community_output <- data.frame(lsoda(y = init_conds, func = SIR_community_model,times = times, parms=parms))

Community_output %>% pivot_longer(cols = !time) %>% arrange(desc(time))%>% 
  ggplot(aes(x=time,y =value, color = name))+geom_line()+
  scale_color_manual(name=NULL,values=c("blue","red","purple"),breaks = c("S","I","R"))+
  theme_classic()+
  labs(x= "Time (days)", y = "Number of individuals")+ggtitle("Community outbreak of pathogen")

```

**Make building**

```{r}

small_bld_3_rooms <- matrix(c(c(1,1,1),
                              c(1,1,1),
                              c(1,1,1)),nrow = 3,ncol = 3)
small_bld_5_rooms <- matrix(c(c(0,1,1,1,1),
                                    c(1,0,1,1,0),
                                    c(1,1,0,1,0),
                                    c(1,1,1,0,0),
                                    c(1,0,0,0,0)),nrow=5, ncol = 5)
diag(small_bld_5_rooms) <-1

church_adjacency_matrix <- matrix(c(c(0,rep(1,3),rep(0,8),rep(1,4),rep(0,30-16)), #column 1 - main area / Hallway
                                    c(1,0,0,1,rep(0,12),rep(1,4),rep(0,30-20)), # column 2 - Hallway
                                    c(1,0,0,1,rep(0,4),rep(1,4),rep(0,30-12)), # column 3 - Hallway
                                    c(1,1,1,0,1,0,1,1,rep(0,30-8)), #column 4 Main room
                                    c(rep(0,3),1,0,1,0,1,rep(0,17),1,rep(0, 30-26)), # # column 5 Hallway
                                    c(rep(0,4),1,0,1,rep(0,15),rep(1,5),0,1,1), # Column 6 Hallway
                                    c(rep(0,3),1,0,1,0,1,rep(0,30-8)), # Column 7 Hallway
                                    c(rep(0,3),1,1,0,1,rep(0,21-8),1,1,rep(0,30-22)), # Column 8 Hallway
                                    rep(c(0,0,1,rep(0,30-3)),4), # columns 9-12
                                    rep(c(1,rep(0,30-1)),4), # columns 13-16
                                    rep(c(0,1,rep(0,30-2)),4), # columns 17-20
                                    rep(c(rep(0,7),1,rep(0, 30-8)),2), # Columns 21 and 22
                                    rep(c(rep(0,5),1,rep(0,30-6)),3), # columns 23-25
                                    c(rep(0,4),1,1,rep(0,30-6)), # Column 26
                                    c(rep(0,5),1,rep(0,21),1,rep(0,30-28)), # column 27
                                    c(rep(0,26),1,rep(0,30-27)), # column 28
                                    c(rep(0,5),1,rep(0,30-6)), # column 29 
                                    c(rep(0,5),1,1,rep(0,30-7))# column 30
                                    
),nrow=30, ncol = 30)
diag(church_adjacency_matrix) <- 1

small_bld_3_rooms_graph<- graph_from_adjacency_matrix(small_bld_3_rooms, mode = "undirected")
#take a look
plot(small_bld_3_rooms_graph)

church_graph <- graph_from_adjacency_matrix(church_adjacency_matrix, mode = "undirected")

```


```{r set carrying capacities}

small_bld_3_rooms_C <- c(100,200,300)

Church_C <- c(20, #column 1 - main area / Hallway
              5, # column 2 - Hallway
              5, # column 3 - Hallway
              362, #column 4 Main room
              5, # # column 5 Hallway
              7, # Column 6 Hallway - slightly bigger hallway
              5, # Column 7 Hallway
              5, # Column 8 Hallway
              50, # Column 9 Large classroom
              30, # column 10 classroom
              30, #column 11 classroom
              2, # column 12 janitors closet
              30, # column 13 classroom
              5, # column 14 - bathroom
              30, # column 15 classroom
              5, # column 16 - bathroom
              30, # column 17 classroom
              2, # column 18 - closet
              30, # column 19 classroom
              50, # column 20 large classroom
              2, # column 21 dressing room
              2, # column 22 dressing room
              4, # column 23 office
              4, # column 24 office
              4, # column 25 office
              4, # column 26 office
              5, # column 27 office - large
              2, # column 28 small bathroom
              4, # column 29 office
              4 # column 30 office
)

length(Church_C)
```
Now that we have building defined by the adjacency matrix, we need to specify the conditions in the building
```{r Building paramaters}
Max_Building_Capacity_small_bld <- sum(small_bld_3_rooms_C) 
Prop_full <- 0.8 # how full do we want our building capacity to be
Adj_Max_Building_Capacity_small_bld <- round(Prop_full*Max_Building_Capacity_small_bld,0) #Adjusted capacity - number of individuals that will be in the building such that building is at 'Prop_full' capacity
delt_small_bld <- Adj_Max_Building_Capacity_small_bld/community_pop # proportionality constant ( what proportion of individuals from the community are in the building of interest)

```

Next, we need to put people in our building. These will be our initial conditions for the building level model.


```{r Not over capacity setup}

day <- 31 #What day to extract results from the community level model
#N_rooms <- length(Church_C)
# delt is the proportionality constant
# C_x is the room capacity for each room
# N_total is the total number of individuals that we want in the building
Bld_setup_func_v2 <- function(Community_output,day, delt,N_rooms, C_x, N_total){
  Building_ICs <- Community_output[day,] #Retrieves the number of S, I, and R individuals in the community at a particular day.
  
  N_x  <- c(rep(0, N_rooms)) #initialize a vector
  for (i in 1:length(C_x)) {
    if (sum(N_x) >= N_total) break  # Stop if all individuals are distributed
    
    # Maximum we can place in this room, constrained by remaining individuals and room capacity
    max_for_room <- min(C_x[i], N_total - sum(N_x))
    
    # Randomly assign a number between 0 and max_for_room
    N_x[i] <- sample(0:max_for_room, 1)
  }
  
  # If after the first pass, there are still individuals left, distribute the remaining in a while loop
  while (sum(N_x) < N_total) {
    # Randomly select a room index
    room_index <- sample(1:length(N_x), 1)
    
    # Check if the selected room can accommodate more individuals
    if (N_x[room_index] < C_x[room_index] && sum(N_x) < N_total) {
      N_x[room_index] <- N_x[room_index] + 1
    }
    
    # Break if we've distributed all individuals or hit the building limit
    if (sum(N_x) >= N_total || sum(N_x) >= N_total) {
      break
    }
  }
  
  #Now we need to find how many S, I, and R individuals we should have in the building based on the proportionality constant solved for above
  Sb <- Community_output$S[day]*delt # number of susceptible in building
  Ib <- Community_output$I[day]*delt # number of Infectious in building
  Rb <- Community_output$R[day]*delt # number of Recovered in building
  
  # calculate the proportion of S,I, and R that should be in the building
  Sb_prop <- Sb/N_total
  Ib_prop <- Ib/N_total
  Rb_prop <- Rb/N_total
  

  # we start with no particles in the building
  P_x <- c(rep(0,N_rooms))
  
  return(data.frame(Sb = Sb, Ib = Ib, Rb = Rb,Sb_prop = Sb_prop,Ib_prop = Ib_prop,Rb_prop= Ib_prop ,P=P_x, N_x = N_x))
}

Small_bld_setup <- Bld_setup_func_v2(Community_output = Community_output,day = day,delt = delt_small_bld,N_rooms =length(small_bld_3_rooms_C),C_x = small_bld_3_rooms_C,N_total = Adj_Max_Building_Capacity_small_bld)
Small_bld_setup

```


**Checking initial conditions**
Now lets check that our building setup function did what I wanted it to. 
Each room's proportion should sum to 1. And if we convert back to numbers using N_x the numbers should equal
```{r props sum to 1}
#Small_bld_setup %>% summarise(total_prop = sum(S)+sum(I)+sum(R),build_pop = sum(N_x))
```
props sum to 1 so that is good

```{r props in building match community}
#prop of susceptible in building
# sum(Small_bld_setup$S)
# #prop of susceptible in community
# Community_output$S[day]/(Community_output$S[day]+Community_output$I[day]+Community_output$R[day])
# 
# #prop of infectious in building
# sum(Small_bld_setup$I)
# #prop of infectious in community
# Community_output$I[day]/(Community_output$S[day]+Community_output$I[day]+Community_output$R[day])
# 
# #prop of recovered in building
# sum(Small_bld_setup$R)
# #prop of susceptible in community
# Community_output$R[day]/(Community_output$S[day]+Community_output$I[day]+Community_output$R[day])
# 

```
Okay, great all of that checks out

**Now people and particles need to move (Transition matrices)**

```{r}
Create_T_Matrix <-function(adjacency_matrix_to_use, N_rooms){
  set.seed(123145) # <- easier for debugging
  T_mov <- data.frame(matrix(runif(N_rooms^2), nrow = N_rooms)) #populates a square matrix/dataframe with random numbers between 0 and 1 for the number of rooms that our building has.
  
  T_mov <- adjacency_matrix_to_use*T_mov #restrict the movement according to our network/adjacency matrix
  T_mov_norm <- t(apply(T_mov, 1, function(x) x / sum(x))) # normalize so that there aren't more people moving than what can (rows should sum to 1)
  T_mov <-T_mov_norm
  return(T_mov)
}

#Church_T_mov <- Create_T_Matrix(adjacency_matrix_to_use = church_adjacency_matrix, N_rooms = length())
Small_bld_T_mov <- Create_T_Matrix(adjacency_matrix_to_use = small_bld_3_rooms,N_rooms = length(small_bld_3_rooms_C))

sum(Small_bld_T_mov[1,])
sum(Small_bld_T_mov[2,])
sum(Small_bld_T_mov[3,])
# sum(Church_T_mov[4,])
# sum(Church_T_mov[5,])
r1_in <- sum(Small_bld_T_mov[,1])
r1_out <- sum(Small_bld_T_mov[1,])
r1_change <- r1_in-r1_out
r2_in <- sum(Small_bld_T_mov[,2])
r2_out <- sum(Small_bld_T_mov[2,])
r2_change <- r2_in-r2_out
r3_in <- sum(Small_bld_T_mov[,3])
r3_out <- sum(Small_bld_T_mov[3,])
r3_change <- r3_in - r3_out
# We can use the same function for the particle matrix
#Church_theta_mov <- Create_T_Matrix(adjacency_matrix_to_use = church_adjacency_matrix)
Small_bld_theta_mov <- Create_T_Matrix(adjacency_matrix_to_use = small_bld_3_rooms,N_rooms = length(small_bld_3_rooms_C))

```

**Now set parameters for the model**


```{r parameters}
parms <-data.frame(s=100,a=5, d=3,lam = 1)
# s = shedding, a = absorption, d = decay, lam = scalar for room capacities
#N_b <- Adj_Max_Building_Capacity #Building population size
Maxtime <- 24*3
times <- seq(from = 0, to = Maxtime, by = 0.2)
m <- 5 # number of equations per room
```

For the model we are going to have to sum over some of the vectors/matrices so we will put those steps in functions. This is needed for the movement of individuals and particles. Particles and people will have to have a different function because there are capacities on rooms for people but not particles.

```{r movement functions}

flux_in_people <- function(N_rooms, Transition_matrix,State,Room_pops,Carrying_capacity,t){
  # Transition_matrix <- matrix(c(0,0.5,1,0.7,0,0,0.3,0.5,0), nrow = 3,ncol = 3)
  # State <- Church_setup$S_prop
  # Room_pops <- Church_setup$N_x
  # Carrying_capacity <- Church_C
  all_room_change <- c(seq(N_rooms))
  for(x in 1:N_rooms){ #flow in to room x from other rooms (j)
    flux_in_temp <- 0
    for(j in 1:N_rooms){
      flux_in_temp <- flux_in_temp + State[j]*Transition_matrix[j,x]*(1-(Room_pops[x]/Carrying_capacity[x]))
    }
    all_room_change[x] <- flux_in_temp
  }
  
  test <- c(M = as.vector(all_room_change))
  return(test)
}

flux_out_people <- function(N_rooms, Transition_matrix,State,Room_pops,Carrying_capacity){
  all_room_change <- c(seq(N_rooms))
  for(x in 1:N_rooms){
    flux_out_temp <- 0
    for(j in 1:N_rooms){
      flux_out_temp <- flux_out_temp + State[x]*Transition_matrix[x,j]*(1-(Room_pops[j]/Carrying_capacity[j]))
    }
    all_room_change[x] <- flux_out_temp
  }
  test <- c(M = as.vector(all_room_change))
  return(test)
}

flux_in_particles <- function(N_rooms, Transition_matrix,State){
  all_room_change <- c(seq(N_rooms))
  for(x in 1:N_rooms){
    flux_in_temp <- 0
    for(j in 1:N_rooms){
      flux_in_temp <- flux_in_temp + State[j]*Transition_matrix[j,x]
    }
    all_room_change[x] <- flux_in_temp
  }
  
  test <- c(M = as.vector(all_room_change))
  return(test)
}

flux_out_particles <- function(N_rooms, Transition_matrix,State){
  all_room_change <- c(seq(N_rooms))
  for(x in 1:N_rooms){
    flux_out_temp <- 0
    for(i in 1:N_rooms){
      flux_out_temp <- flux_out_temp + State[x]*Transition_matrix[x,i]
    }
    all_room_change[x] <- flux_out_temp
  }
  test <- c(M = as.vector(all_room_change))
  return(test)
}

```


**Function for the whole model**

```{r sanity check}
Particle_model_v3 <- function(t, x, parms,T_mov, theta_mov, adjacency_matrix_to_use,C_x,N_b,N_rooms,Ib_prop){
  # x <- Church_Init_conds
  # T_mov <- Church_T_mov
  # theta_mov <- Church_theta_mov
  # adjacency_matrix_to_use <- small_bld_3_rooms
  # C_x <- small_bld_3_rooms_C
  # N_b <- N_b
  N_rooms <- N_rooms
  ncompartment <- 2
  n_rooms <- length(x)/ncompartment
  N_x <- as.matrix(x[1:n_rooms])
  P <- as.matrix(x[(n_rooms+1):(2*n_rooms)])
  
  with(parms,{
    
    dN_x <- as.matrix((flux_in_people(N_rooms=N_rooms, Transition_matrix =T_mov, State=N_x,Room_pops = N_x,Carrying_capacity = C_x)) - flux_out_people(N_rooms=N_rooms, Transition_matrix = T_mov, State=N_x,Room_pops = N_x,Carrying_capacity = C_x))
    
    dP <- s*Ib_prop*as.matrix(N_x) - as.matrix(a*P/(lam*C_x*as.matrix(N_x)))-d*as.matrix(P) + as.matrix(as.matrix(flux_in_particles(N_rooms=N_rooms, theta_mov,State = P)) - as.matrix(flux_out_particles(N_rooms=N_rooms, theta_mov,State = P)))
    
    # 
    dt <- c(dN_x,dP)
    
    return(list(dt))})
  
}
```



```{r}
#Church_Init_conds_v1 <-c(S=Church_setup$S, I = Church_setup$I, R = Church_setup$R, P = Church_setup$P)

Small_bld_Init_conds_v3 <-c(N_x =Small_bld_setup$N_x, P = Small_bld_setup$P)


Small_bld_output_v3 <- data.frame(lsoda(y = Small_bld_Init_conds_v3, func = Particle_model_v3,times = times,
                                  parms = parms,
                                  adjacency_matrix_to_use=small_bld_3_rooms,
                                  theta_mov =Small_bld_theta_mov,
                                  T_mov = Small_bld_T_mov, 
                                  C_x=small_bld_3_rooms_C,
                                  N_b = Adj_Max_Building_Capacity_small_bld,
                                  N_rooms = length(small_bld_3_rooms_C),
                                  Ib_prop = Small_bld_setup$Ib_prop[1]))

```


```{r}
data_clean <- Small_bld_output_v3%>% pivot_longer(cols = !time,
                                                   names_to = c("State", "Room"),
                                                   names_pattern = "(P|N_x)(\\d+)",
                                                   values_to = "Number")


data_clean %>% filter(State == "P") %>% 
  group_by(State,Room)%>% 
  ggplot( aes(x=time, y = Number, group =Room, color = Room))+
  geom_line()+theme_classic()+
  labs(x = "Time (hours)", y= "Proportion of Susceptible individuals")+ggtitle("Proportion of Susceptible individuals across rooms")

data_clean %>% filter(State == "N_x") %>% 
  group_by(State,Room)%>% 
  ggplot( aes(x=time, y = Number, group =Room, color = Room))+
  geom_line()+theme_classic()+
  labs(x = "Time (hours)", y= "Proportion of Infectious individuals")+ggtitle("Proportion of Infectious individuals across rooms")

data_clean %>% filter(State == "N_x") %>% ungroup() %>% group_by(time) %>% summarise(Total_prop=sum(Number))%>%
  ggplot(aes(x=time, y=Total_prop))+geom_line()+
  labs(x = "Time (hours)", y= "Proportion of individuals")+ggtitle("Proportion of individuals in rooms within a building")
temp <- data_clean
temp_C_x <- data.frame(Room = as.character(seq(1:length(small_bld_3_rooms_C))),C_x = c(small_bld_3_rooms_C))
temp_data <- left_join(temp,temp_C_x, by = "Room")

ratio_data <- temp_data %>% pivot_wider(names_from = c(State),values_from = c(Number)) %>% 
   group_by(time, Room)%>%mutate(s =(parms$s),a = parms$a,lam = parms$lam, Ib_prop = Small_bld_setup$Ib_prop[1],Numerator = s*N_x*Ib_prop*C_x*lam , Denominator =(a*P), ratio = Numerator/Denominator ) 

#(parms$s*N_x*Small_bld_setup$Ib_prop[1]*N_x*small_bld_3_rooms_C[Room]*parms$lam)
ggplot(ratio_data, aes(x = time, y = ratio, color = Room))+geom_line()
```


```{r}
p1_v2
p2_v2
p3_v2
p4_v2
p5_v2
p6_v2
p7_v2
p8_v2

```




